Cybersecurity impacts at the beginning\\ of Covid-19 pandemic in Italy

### 2021-03-08 article.tex
Minor fix

The advent of Covid-19 pandemic and the consequent lock-downs have led companies to face needs such as smart working, remotely work and digital business,
accelerating their digitalization efforts. As reported in \cite{Goasduff20} by Gartner, a world's leading research and advisory company, before the pandemic crisis
most organizations moved their digital strategies forward at a steady pace, but during the Covid-19 pandemic there has been a significant increase in the 
development of digital products and services to maintain and accelerate customer engagement. However, the digitalization race has led to many cyber security 
issues and the intensification of malicious cyber attacks around the world~\cite{HKICS20,PA21}. In this scenario, Italy is an interesting case of study because 
it is composed mostly from Small Medium Enterprises (SMEs) and therefore can highlight specific aspects instead of the various global views in which large companies 
have a greater influence. This paper presents a window of italian panorama to understand how the cyber-security is faced up during COVID-19 pandemic. 
Section~\ref{sec:DigitalAttacksObservatory} presents \textit{Digital Attacks Observatory}, a survey containing relevant information about cyber security and 
italian organizations which will be used in subsequent sections as a reference to better understand Covid-19 impacts. A detailed discussions of collected data are in 
Sections~\ref{sec:DataDiscussion}, \ref{sec:TypeAttacks}. Finally, Section~\ref{sec:Conclusion} concludes by underlying the importance of cybersecurity culture focusing 
on Italian reality and SMEs.

@@

The \textit{Digital Attacks Observatory} (OAD)~\cite{oadweb} is the only independent online survey in Italy about intentional digital attacks on IT systems of companies and public
bodies operating and in addition drawn up with the precious collaboration of the Italian Postal and Telecommunications Police. The OAD survey does not provide 
for a predefined set of respondents, but it allows potential interested parties full and free access to an online questionnaire, in a totally anonymous manner. 
Thanks to the number of responses collected and their balanced distribution between companies and public bodies of various sizes and belonging to various product 
sectors, the OAD survey provides a specific picture of the cyber attacks in Italy.

@@

OAD 2020 survey\footnote{The report is written in Italian, only the Executive Summary is in English: 186 A4 pages, 148 images and graphics, 11 Chapters
(147 A4 pages) and 9 Attachments (39 A4 pages)}~\cite{oad20} took place during the Covid-19 pandemic, and it covered the entire 2019 and the first quarter 
of 2020, when in Italy the pandemic exploded. The data comparison of 2019 and the first quarter 2020, provided by the same respondents pool, has a statistical relevance and provides a 
clear indication of the Covid-19 pandemic on the intentional cyber attacks in Italy. This pandemic has been the trigger for a wide range of cyber attacks, mainly 
caused by the sudden - and in part unprepared - passage of many to agile working from home and to a strong use of every type of IT service on Internet, mainly
derived by the mobility lockdowns imposed by the Italian Authorities. The emerged respondents’ pool covers all the product sectors, including Public Administrations,
even if the majority of the respondents’ companies belong to the ICT sector (30.8\%). The 2020 pool is well balanced for the size of the organizations, in terms
of employees, between those below 250 and the largest ones. It must be taken into account that in Italy 99.91\% \cite{istat21}
of enterprises are SMEs, under 250 employees, and 95\% are under ten employees. The OAD 2020 survey is therefore able to consider small and 
tiny organizations, which are thee vast majority in Italy, and which are not normally considered in other cyber security surveys: 57.5\% of the OAD 2020 respondents
belong to structures with less than 250 employees, and of these 22.1\% under 10.

@@

\item As pointed out before, 99,91\% of the companies in Italy are SME, and 95\% are under 10 employees. Specifically in the 2020 OAD survey, the respondents are 57,7\% for 
SME/small P, and of these 22,1\% are under 10 employees. In general, cybercriminals do not waste time in attacks targeting small businesses, to obtain only a marginal profit, 
with the risk of being caught by the police.
\end{itemize}

@@

Fig.~\ref{fig:3} confirms what already underlined before: the attackers try to hit the big structures, because with these it is more probable the realization of gains, even if illegal.
The comparison between the whole 2019 and the first quarter 2020 maintains the same logical trend. The only meaningful difference is for the frequent attacks relieved by the very
small organizations, up to 10 employees: 0\% in 2019 grows up to 22,2\% (see in Fig.~\ref{fig:3} the red line). Such a strong increase derives by the large utilization of e-baking and 
e-commerce services, with the related cyber attacks.

@@

\section{Type of attacks}\label{sec:TypeAttacks}

@@

\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{pictures/fig5.png}
		\caption{}
		\label{fig:5}
\end{figure}
\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{pictures/fig6.png}
		\caption{}
		\label{fig:6}
\end{figure}

As pointed out in Fig.~\ref{fig:1}, the Informatics Systems considered in the 2020 survey and their cybersecurity  are in the medium-high range and the emerged information show a relevant 
improvement in both technical and organizational measures, in comparison to the previous surveys. Few of the considered Italian Informatics Systems are based on Data Centers in 
Italy, most of the respondents’ companies have medium-small Informatic Systems partly on premise and partly outsourced, with an increasing use of cloud services.
The high number of attacks detected in 2018, and the privacy obligations, related to General Data Protection Regulation (GDPR), have certainly contributed to strengthening cyber security measures, and a further improvement 
comes from the growing use of cloud services, where usually there are high standard security measures. Organizational measures for cybersecurity, historically lacking and neglected
in Italy, have improved in terms of defining roles and separation of duties, of organizational policies and procedures, and of incidents management. These measures often are lacking
in small organizations, and in general the cybersecurity awareness and competence are low, and mainly at the top level of the public and private organizations. In Italy, there is 
still a long way to go in terms of cybersecurity continual training and awareness. A formal and bureaucratic approach for the organizational procedures is still prevalent, which 
once defined, often do not find a possible concrete applicability, periodic updates and operational tests: an emblematic example is represented by the Disaster Recovery plans for 
which, several times, the required alternative ICT resources to be used are not forecasted and allocated and there also not provided periodic exercises and simulations of the 
possible disasters. Cybersecurity management tools still have limited diffusion among the 2020 OAD respondents, in particular the most advanced ones based on artificial intelligence
techniques. Although the use of IoT \cite{JK21}, industrial automation, robotics systems \cite{SGLMDXHKCZT21}, and systems based on blockchain technologies \cite{RMFF21} are expanding and can allow to combat the Covid-19 pandemic, in 
the OAD 2020, we can find only a low number of respondents involved, which also derives from the limited percentage of their product sectors that should be the more interested in these systems: the manufacturing sectors, the logistics, the research and 
development centers and labs, the Local Public Administrations for territorial control, and so on. For these topics, the data that emerged from the survey are considered only as 
specific cases that contribute to the general values on the cyberattacks, but which cannot be still considered of reference at the Italian level.

@@

The OAD 2020 survey highlights a situation of cyberattacks of various types but all technically of ian high complexity and sophistication, with a slightly increasing 
spread  in Italy  compared to the trend that emerged in the twelve years of OAD surveys, but lower than the peak of 2018. Despite the proliferation of cyberattacks to varying 
degrees related to the Covid-19 pandemic, in the first four months of 2020 the general spread of attacks is at similar values to those of 2019: to be verified with the next OAD 2021
if there will be changes to this trend. The Italian reality, made up of a very large number of small and tiny organizations, does not make our country one of the most attractive for cybercriminals, but cyber warfare
and massive attacks represent a growing and serious risk, as has already happened in part with the widespread of ransomware on computer systems whit a lack of or  with low 
cybersecurity  basic measures. OAD 2020 notes a clear improvement and strengthening of digital security measures, both technical and organizational, even if the most modern 
prevention, protection and management systems that use artificial intelligence techniques are still embryonic  in the respondents' basin. The defense measures and techniques 
in use chase the increasingly sophisticated and smart evolution of attacks, but almost always late. The high density of vulnerabilities 
requires different approaches and new logics, with the aim of making all ICT systems interconnectable to the Internet intrinsically safe, by default and by design. But we are 
still a long way from this goal, and in order to decisively improve the concrete fight against continuous attacks and cybercrime, it is currently necessary to increase 
cybersecurity’ awareness and skills at all levels, an effective collaboration between the police at world level, and primarily a real and large usage of the professional ethics
 both of those involved (supply side) and both of those who decide (demand side) on cybersecurity.

>>>>

The advent of Covid-19 pandemic and the consequent lock-downs have led companies to face needs such as smart working, remotely work and digital business,
accelerating their digitalization efforts. As reported in \cite{Goasduff20} by Gartner, a world's leading research and advisory company, before the pandemic crisis
most organizations moved their digital strategies forward at a steady pace, but during the Covid-19 pandemic there has been a significant increase in the 
development of digital products and services to maintain and accelerate customer engagement. However, the digitalization race has led to many cyber security 
issues and the intensification of malicious cyber attacks around the world~\cite{HKICS20,PA21}. In this scenario, Italy is an interesting case of study because 
it is composed mostly from Small Medium Enterprises (SMEs) of different sectors and therefore can highlight specific aspects instead of the various global views 
in which large companies have a greater influence. This paper presents a window of italian panorama to understand how the cyber-security is faced up during COVID-19 pandemic. 
Section~\ref{sec:DigitalAttacksObservatory} presents \textit{Digital Attacks Observatory}, a survey where we have collected relevant information about cyber security and 
italian organizations which will be used in subsequent sections as a reference to better understand Covid-19 impacts. A detailed discussions of collected data and trends are in 
Sections~\ref{sec:DataDiscussion}, \ref{sec:TypeAttacks}. Finally, Section~\ref{sec:Conclusion} concludes by underlying the importance of cybersecurity culture focusing 
on Italian reality.

@@

The \textit{Digital Attacks Observatory} (OAD)~\cite{oadweb} is the only independent online survey in Italy about intentional digital attacks on IT systems of companies and public
bodies operating and in addition drawn up with the precious collaboration of the Italian Postal and Telecommunications Police. The OAD survey does not provide 
for a predefined set of respondents, but it allows potential interested parties full and free access to an online questionnaire, in a totally anonymous manner. 
Thanks to the number of responses collected and their balanced distribution between companies and public bodies of various sizes and belonging to various product 
sectors, we can provide a specific picture of the cyber attacks in Italy.  The last OAD 2020  reaches the twelfth year of consecutive surveys on cyberattacks in Italy.

@@

OAD 2020 survey\footnote{The report is written in Italian, only the Executive Summary is in English: 186 A4 pages, 148 images and graphics, 11 Chapters
(147 A4 pages) and 9 Attachments (39 A4 pages)}~\cite{oad20} took place during the Covid-19 pandemic, and it covered the entire 2019 and the first quarter 
of 2020, when in Italy the pandemic exploded. The data comparison of 2019 and the first quarter 2020, provided by the same respondents pool, has a statistical relevance and provides a 
clear indication of the Covid-19 pandemic on the intentional cyber attacks in Italy. This pandemic has been the trigger for a wide range of cyber attacks, mainly 
caused by the sudden - and in part unprepared - passage of many to agile working from home and to a strong use of every type of IT service on Internet, mainly
derived by the mobility lockdowns imposed by the Italian Authorities. The emerged respondents’ pool covers all the product sectors, including Public Administrations,
even if the majority of the respondents’ companies belong to the ICT sector (30.8\%). The 2020 pool is well balanced for the size of the organizations, in terms
of employees, between those below 250 and the largest ones. It must be taken into account, as reported by ISTAT~\cite{istat21}, that in Italy 99.91\%
of enterprises are SMEs, under 250 employees, and 95\% are under ten employees. The OAD 2020 survey is therefore able to consider small and 
tiny organizations, which are thee vast majority in Italy, and which are not normally considered in other cyber security surveys: 57.5\% of the OAD 2020 respondents
belong to structures with less than 250 employees, and of these 22.1\% under 10.

@@

\item In general, cybercriminals do not waste time in attacks targeting small businesses, to obtain only a marginal profit, 
with the risk of being caught by the police.  As confirmed by trends of Fig.~\ref{fig:3}, the attackers try to hit the big structures,
 because with these it is more probable the realization of gains, even if illegal.
\end{itemize}

@@

As already highlighted above, the comparison between the whole 2019 and the first quarter 2020 in Fig.~\ref{fig:3} maintains the same logical trend. The only meaningful difference is for the frequent attacks relieved by the very
small organizations, up to 10 employees: 0\% in 2019 grows up to 22,2\% (see in Fig.~\ref{fig:3} the red line). Such a strong increase derives by the large utilization of e-baking and 
e-commerce services, with the related cyber attacks.

@@

\section{Cyber attack trends}\label{sec:TypeAttacks}

@@

\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{pictures/fig5.png}
		\caption{Diffusion of digital attacks on respondents' computer systems based on the type of attack between 2019 - 1st quoter 2020 (multiple answers) \cite{oad20}}
		\label{fig:5}
\end{figure}
\begin{figure}
	\centering
		\includegraphics[width=0.90\textwidth]{pictures/fig6.png}
		\caption{Average diffusion (percentage) of the attack techniques detected in the most serious attacks between 2019 - 1st quoter 2020 (multiple answers) \cite{oad20}}
		\label{fig:6}
\end{figure}

As pointed out in Fig.~\ref{fig:1}, the Informatics Systems considered in the 2020 survey and their cybersecurity  are in the medium-high range and the collected information show a relevant 
improvement in both technical and organizational measures, in comparison to the previous surveys. Few of the considered Italian Informatics Systems are based on Data Centers in 
Italy, most of the respondents’ companies have medium-small Informatic Systems partly on premise and partly outsourced, with an increasing use of cloud services.
The high number of attacks detected in 2018, and the privacy obligations, related to General Data Protection Regulation (GDPR), have certainly contributed to strengthening cyber security measures, and a further improvement 
comes from the growing use of cloud services, where usually there are high standard security measures. Organizational measures for cybersecurity, historically lacking and neglected
in Italy, have improved in terms of defining roles and separation of duties, of organizational policies and procedures, and of incidents management. These measures often are lacking
in small organizations, and in general the cybersecurity awareness and competence are low, and mainly at the top level of the public and private organizations. In Italy, there is 
still a long way to go in terms of cybersecurity continual training and awareness. A formal and bureaucratic approach for the organizational procedures is still prevalent, which 
once defined, often do not find a possible concrete applicability, periodic updates and operational tests: an emblematic example is represented by the Disaster Recovery plans for 
which, several times, the required alternative ICT resources to be used are not forecasted and allocated and there also not provided periodic exercises and simulations of the 
possible disasters. Cybersecurity management tools still have limited diffusion among the 2020 OAD respondents, in particular the most advanced ones based on artificial intelligence
techniques. Although the use of IoT \cite{JK21}, industrial automation, robotics systems \cite{SGLMDXHKCZT21}, and systems based on blockchain technologies \cite{RMFF21} are expanding and can allow to combat the Covid-19 pandemic, in 
the OAD 2020, we can find only a low number of respondents involved, which also derives from the limited percentage of their product sectors that should be the more interested in these systems: the manufacturing sectors, the logistics, the research and 
development centers and labs, the Local Public Administrations for territorial control, and so on. For these topics, the data that emerged from the survey are considered only as 
specific cases that contribute to the general values on the cyberattacks, but which cannot be still considered of reference at the Italian level.

@@

The OAD 2020 survey highlights a situation of cyberattacks of various types but all technically of an high complexity and sophistication, with a slightly increasing 
spread  in Italy  compared to the trend that emerged in the twelve years of OAD surveys, but lower than the peak of 2018. Despite the proliferation of cyberattacks to varying 
degrees related to the Covid-19 pandemic, in the first four months of 2020 the general spread of attacks is at similar values to those of 2019: to be verified with the next OAD 2021
if there will be changes to this trend. The Italian reality, made up of a very large number of small and tiny organizations, does not make our country one of the most attractive for cybercriminals, but cyber warfare
and massive attacks represent a growing and serious risk, as has already happened in part with the widespread of ransomware on computer systems whit a lack of or  with low 
cybersecurity  basic measures. OAD 2020 notes a clear improvement and strengthening of digital security measures, both technical and organizational, even if the most modern 
prevention, protection and management systems that use artificial intelligence techniques are still embryonic  in the respondents' basin. The defense measures and techniques 
in use chase the increasingly sophisticated and smart evolution of attacks, but almost always late. The high density of vulnerabilities 
requires different approaches and new logics, with the aim of making all ICT systems interconnectable to the Internet intrinsically safe, by default and by design. But we are 
still a long way from this goal, and in order to decisively improve the concrete fight against continuous attacks and cybercrime, it is currently necessary to increase 
cybersecurity’ awareness and skills at all levels, an effective collaboration between the police at world level, and primarily a real and large usage of the professional ethics
 both of those involved (supply side) and both of those who decide (demand side) on cybersecurity.
